Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-07 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1986 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 664c83d3f02ebb28f15d4c70a6c08764667c492c
revised large parts of intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1988 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e3bfcc5ad9c4a6e7c4106b2334bf896610df77c1

\SweaveOpts{eval = TRUE}
\section{Introduction}
In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates. Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has led to the development of generalized additive models (GAM) also capturing non-linear effects. In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate. In the left picture the expected value of the response variable is modeled by a linear function in the covariate. In the middle a GAM is applied where a smooth function allows for more flexibility.

These models consider only the expectation of the underlying distribution. However, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting. For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates. Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation. Therefore, these models provide common and widely used tools, also due to their intuitive structure and interpretability. However, the restriction to model only linear effects has risen the interest in applying more flexible models. Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM as illustrated in the right plot of Figure~\ref{Devel_parmod}. These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g. areas where no abrupt changes in meteorological/climatological events appear. However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model. In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{breiman1984classification})
%%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
%(\citealt{loh2011classification}) 
or random forests such as Breiman and Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process. Therefore, no additional effort or knowledge for the specification of the model is required in advance. However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well. In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.

@@

@@

@@

The exact way of determining how and where to split varies across different tree algorithms. For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix. Summed up, the steps of building a distributional tree are:
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \theta) = 0$
\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose the breakpoint with the highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is reached.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}
The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.
%%% <TH> we need a little more detail here or a reference to a better
%%% description of mob/ctree OR appendix</TH>

@@

Forest models differ mainly in the way how the trees are combined to make predictions. Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process. In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations. This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``bagging survival trees'' by \cite{hothorn2004}. The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (“bagging survival trees”, Hothorn et al. 2004). 
In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
$$
w^{\text{forest}}_i(\bold{z}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
$$
where $T$ is the number of trees of the forest, $B^t$ is the number of terminal nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th terminal node of the \textit{t}-th tree.

>>>>

\SweaveOpts{eval = TRUE}
\section{Introduction}

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1986})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

While in many applications conditional mean regression models had been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
that there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear
models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogenous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
considers both the mean as well as the uncertainty of the NWP model outputs
to obtain probabilistic forecasts calibrated to a particular site.
%% late: introduce MOS and EMOS? more details about precipitation forecasting

@@

However, for example in complex
terrain in mountainous regions various factors can change rapidly and their
effects are difficult to be represented by a smooth parametric model. In this
situation a different approach which can deal with these abrupt changes would be
to apply tree-structured models such as regression trees
(\citealt{breiman1984classification}) or random forests such as Breiman and
Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance. However, even though their
main advantage of being able to model non-additive effects and interactions
without any necessary prespecification can play to their strength, regression
trees are not an adequate choice if (additional) smooth effects appear as well.
In this situation a random forest provides a good compromise as it combines a
set of trees such that the steps imposed by the trees are smoothed out in case
of smooth effects while abrupt changes are still captured by the tree structure
of the model.

@@

The exact way of determining how and where to split varies across different tree algorithms. For distributional trees one possible choice is a permutation test based algorithm which is explained in detail in the appendix. Summed up, the steps of building a distributional tree are:
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\bold{\hat{\theta}}$ by solving $\sum_{i=1}^n s(y_i, \hat{\theta}) = 0$.
\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose the breakpoint with the highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is reached.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}
The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.
%%% <TH> we need a little more detail here or a reference to a better
%%% description of mob/ctree OR appendix</TH>

@@

Forest models differ mainly in the way how the trees are combined to make predictions. Following the strategy explained and proposed by \cite{transformationforests2017} we interpret random forests as adaptive local likelihood estimators as distributional models are fit locally in subgroups while the term ``adaptive'' refers to the dependence on the splitting variables $\bold{Z}$ and the resulting subgroups and weighing schemes. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process. In transformation forests by \cite{transformationforests2017} so-called ``nearest neighbor weights'' are calculated as a measure of distance between observations. This approach has also already been applied in several random forest-type methods such as in ``adaptive nearest neighbors'' by \cite{lin2006}, in ``quantile regression forests''  by  \cite{meinshausen2006} and in ``bagging survival trees'' by \cite{hothorn2004}. The idea of nearest neighbor weights is to measure similarity of a pair of observations by counting the number of trees in which it is asigned to the same terminal node.
%Similar observations have a high probability of ending up in the same terminal node whereas this probability is low for quite different observations. 
%although the first random forest-type algorithm for the estimation of conditional distribution functions was published more than a decade ago (bagging survival trees, Hothorn et al. 2004). 
In particular, for a new set of covariates $\bold{z} = (z_1, \ldots, z_m) $ such a set of weights 
$\bold{w}^{\text{forest}}~=~(w^{\text{forest}}_1, \ldots, w^{\text{forest}}_n)$ is obtained by the following procedure: For each tuple $(y_i,\bold{z}_i)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_i,\bold{z}_i)$ ends up in the same terminal node as the new set of covariates $\bold{z}$:
$$
w^{\text{forest}}_i(\bold{z}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
$$
where $T$ is the number of trees of the forest, $B^t$ is the number of terminal nodes of the \textit{t}-th tree and $\mathcal{B}^t_b$ represents the \textit{b}-th terminal node of the \textit{t}-th tree.
